### 2016-02-12 doc/thesis/intro.tex
Updated reading and writing

Speciation often occurs by an \defn{allopatric} process, where a sub-population
of organisms is isolated from the original population by a geographic
barrier~\autocite{coyne2004speciation}. Over time, differing selection
pressures on either side of the barrier, coupled with genetic drift, cause the
two populations to diverge genetically. This eventually results in two distinct
species, which will appear as two subtrees or \defn{clades} in their species
tree if descendants of both survive to the present. Speciation is not
necessarily allopatric: \defn{sympatric} speciation refers to the emergence of
a new species without any geographic barrier~\autocite{coyne2004speciation}.
Indeed, there is a whole continuum of possible speciation
processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
completely sympatric. However, at least for eukaryotic organisms, allopatric
speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

%Of course, this is not
%necessarily the case, as selection may cause one or both of the new
%sub-populations or their descendants to become extinct. On the other hand,
%selection may favour one of the sub-populations, allowing it to expand into new
%regions or niches and possibly paving the way for subsequent speciation events.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the pathogen
population of the individual was sampled. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).

\textit{A priori}, transmission trees and species trees have nothing to do with
each other. One represents the evolutionary history of a group of organisms,
the other traces the spread of a disease through a host population. However,
when considering epidemics of RNA viruses, transmission trees and species trees
are highly related. The viral population within a single host has often been
referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
the fast evolutionary rate of these viruses, two hosts' viral populations are
invariably genetically distinct. The transmission process is functionally
identical to allopatric speciation, involving geographic isolation of a
sub-population in a new host and subsequent genetic divergence. However, for
the most part, it is not selection but rather host epidemiology which causes
certain lineages to proliferate while driving others to extinction. A host who
recovers, dies, or is isolated equates to an extinction in the transmission
tree. On the other hand, dense populations in frequent contact often experience
rapid epidemic growth, which causes similarly rapid accumulation of branching
points in the tree.

@@

Finally, we must note two important points. First, we have not yet considered
the selection pressures operating within a single host, such as immune
pressure, because these forces do not directly impact transmission trees unless
they cause a host to recover or die. Second, we reiterate that both
transmission trees and species trees are theoretical objects which are almost
never known with certainty for real data. Indeed, the estimation of
transmission trees is often problematic, and it is in such estimation that
within-host forces play a role. Both of these issues are discussed further
below in subsection~\ref{subsubsec:treeconv}, though we continue to ignore them
in the next subsection.

@@

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure assocates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1.
In the context of infectious diseases, the \gls{nltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
kernel is a \defn{similarity measure}, rather than a distance measure, as it is
maximized when the two trees being compared are the same. The basis of the tree
kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibtive. In
the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
to fit epidemiological models to observed trees.

@@

A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
incorproates the phenomenon of \defn{preferential attachment} seen in real life
networks. This means that new connections are more likely to involve nodes
which already have a high number of connections - popular nodes tend to become
more popular. In the formulation we consider here, BA networks have two
parameters, $m$ and $\alpha$, which are used to construct the network via the
following algorithm. The network begins as a single node, then the remaining
$n-1$ nodes are added one at a time. Each time a new node is added, it is
connected to $m$ other nodes. The probability of one of these connections
involving an existing node of degree $k$ is proportional to $k^\alpha$. Note
that in the original formulation~\autocite{barabasi1999emergence}, there was an
additional parameter $m_0$ which indicated the number of initial vertices in
the network, and the only value of $\alpha$ considered was 1. For our purposes,
we fix $m_0 = 1$ but allow $\alpha$ to vary. We refer to $\alpha$ throughout
this work as the \defn{preferential attachment power} or just attachment power.

@@

WS and BA networks differ from ER networks in an important way: it is possible
to explicitly calculate the likelihood of the ER model parameter $p$ given an
observed network, but this cannot be done for the other two models. This is
because the probability of a particular edge occuring is always equal to $p$ in
an ER network, regardless of the presence or absence of any other edges. In
other words, the edges can be viewed as a series of $\binom{n}{2}$ Bernoulli
trials with success probability $p$, so that the likelihood of $p$ given an
observed graph with $k$ edges is distributed as 
Binomial$\left(\binom{n}{2}, p\right)$.

@@

@@

A linear regression is a model of the relationship between some data $\vec{x} =
(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
assumes that the outcomes are linearly related to the data, modulo some noise
introduced by, say, measurement errors and environmental fluctuations. It other
words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\[
  y_i = \beta x_i + \N(0, \sigma^2),
\]
where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
the model. If the $y_i$ are all independent, then for particular choics of
$\beta$ and $\sigma$, we can write down the probability density of observing
$\vec{y}$ given $\vec{x}$.
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
\]
The higher the value of this probability density, the more likely the
observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
natural criterion for choosing the parameters: we want to pick $\beta$ and
$\sigma$ which define a model where the probability of our observed data is as
high as possible. When performing such an optimization for fixed data, the
density function just written is also called the \defn{likelihood} of the
parameters,
\[
  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
\]
The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
\vec{x}, \vec{y})$ are called the \gls{ML} estimates.

@@

The main idea of approximate Bayesian computation is that the ``best''
parameters will define a model which produces simulated data similar to the
real data. Suppose we fix values of $b$ and $\sigma$. Then, using the given
data points $x_i$, we can generate simulated observations $y_i^*$ by adding
$bx_i$ and a randomly chosen error drawn from $\N(0, \sigma)$.

@@

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly}. Effectively, it comes down to the
approach described in the previous subsection of guessing parameter values
until one is close enough to the truth. More specificially, a possible set of
parameters $\theta$ is drawn from the prior distribution, and a simulated data
set $z$ is generated from the model with those parameters. If the distance
between the simulated data set and the real data, $\rho(\eta(y), \eta(z))$, is
small enough, then we accept $\theta$ as a sample from the posterior. This can
be repeated until as many samples as desired are obtained.

>>>>

Speciation often occurs by an \defn{allopatric} process, where a sub-population
of organisms is isolated from the original population by a geographic
barrier~\autocite{coyne2004speciation}. Over time, differing selection
pressures on either side of the barrier, coupled with genetic drift, cause the
two populations to diverge genetically. This eventually results in two distinct
species, which will appear as two subtrees or \defn{clades} in their species
tree if descendants of both survive to the present. Of course, this is not
necessarily the case, as selection may cause one or both of the new
sub-populations or their descendants to become extinct. On the other hand,
selection may favour one of the sub-populations, allowing it to expand into new
regions or niches and possibly paving the way for subsequent speciation events.
Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
the emergence of a new species without any geographic
barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
completely allopatric to completely sympatric. However, at least for eukaryotic
organisms, allopatric speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labelled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the pathogen
population of the individual was sampled. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).

At first glance, transmission trees and species trees seem to have nothing to
do with each other. One represents the evolutionary history of a group of
organisms, the other traces the spread of a disease through a host population.
However, when considering epidemics of RNA viruses, transmission trees and
species trees are highly related. The viral population within a single host has
often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
and due to the fast evolutionary rate of these viruses, two hosts' viral
populations are invariably genetically distinct. The transmission process is
functionally identical to allopatric speciation, involving geographic isolation
of a sub-population in a new host and subsequent genetic divergence. However,
for the most part, it is not selection but rather host epidemiology which
causes certain lineages to proliferate while driving others to extinction. A
host who recovers, dies, or is isolated equates to an extinction in the
transmission tree. On the other hand, dense populations in frequent contact
often experience rapid epidemic growth, which causes similarly rapid
accumulation of branching points in the tree.

@@

Finally, we must note two important points. First, we have not yet considered
the selection pressures operating within a single host, such as the immune
response. Second, we reiterate that both transmission trees and species trees
are theoretical objects which are almost never known with certainty for real
data. Indeed, the estimation of transmission trees is often problematic, and it
is in such estimation that within-host forces play a role. Both of these issues
are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
continue to ignore them in the next subsection.

@@

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure associates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1.
In the context of infectious diseases, the \gls{nltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
kernel is a \defn{similarity measure}, rather than a distance measure, as it is
maximized when the two trees being compared are the same. The basis of the tree
kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibitive. In
the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
to fit epidemiological models to observed trees.

@@

A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
incorporates the phenomenon of \defn{preferential attachment} seen in real life
networks. This means that new connections are more likely to involve nodes
which already have a high number of connections - popular nodes tend to become
more popular. In the formulation we consider here, BA networks have two
parameters, $m$ and $\alpha$, which are used to construct the network via the
following algorithm. The network begins as a single node, then the remaining
$n-1$ nodes are added one at a time. Each time a new node is added, it is
connected to $m$ other nodes. The probability of one of these connections
involving an existing node of degree $k$ is proportional to $k^\alpha$. Note
that in the original formulation~\autocite{barabasi1999emergence}, there was an
additional parameter $m_0$ which indicated the number of initial vertices in
the network, and the only value of $\alpha$ considered was 1. For our purposes,
we fix $m_0 = 1$ but allow $\alpha$ to vary. We refer to $\alpha$ throughout
this work as the \defn{preferential attachment power} or just attachment power.

@@

WS and BA networks differ from ER networks in an important way: it is possible
to explicitly calculate the likelihood of the ER model parameter $p$ given an
observed network, but this cannot be done for the other two models. This is
because the probability of a particular edge occurring is always equal to $p$ in
an ER network, regardless of the presence or absence of any other edges. In
other words, the edges can be viewed as a series of $\binom{n}{2}$ Bernoulli
trials with success probability $p$, so that the likelihood of $p$ given an
observed graph with $k$ edges is distributed as 
Binomial$\left(\binom{n}{2}, p\right)$.

@@

\section{Sequential Monte Carlo}

\Gls{SMC} is a statistical inference method which samples from a sequence of
probability distributions in a fixed order~\autocite{del2006sequential}. This
idea of \defn{sequential sampling} is useful several contexts, but we consider
here its application to model fitting in a Bayesian setting.

\texttt{TODO: move the basics of model fitting (likelihoods and priors) to here}

Our objective is to obtain samples from, and summary statistics of, the
posterior distribution on the model of interest's parameters. \gls{SMC} can be
applied in this context by defining a sequence of distributions, starting from
the posterior and ending at the prior, which constitute a ``smooth''
trajectory. The main \gls{SMC} algorithm for this problem was developed by
\textcite{del2006sequential}, based on an existing methodology called
\gls{SIS}. \gls{SIS} is also designed to sample from a sequence of
distributions, but rather than all being defined on the same space (such as
model parameters), they are defined on nested spaces of increasing dimension.
Following \textcite{del2006sequential} and other reviews of
\gls{SMC}~\autocite{doucet2001introduction}, we will begin this section by
describing \gls{SIS}, and then turn to the adaptation of the method to the case
when the sequence of distributions are all defined on the same space. Note that
\textcite{del2006sequential} use the variable $\pi$ for the target
distributions, but \textcite{doucet2001introduction} use $\pi$ to indicate the
importance distributions. We again follow the former and denote target
distributions by $\pi$ and importance distributions by $\eta$.

The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
statistics of distributions which are known only up to a normalizing constant,
and therefore cannot be sampled from directly. That is, if $\pi$ is such a
distribution and $f$ is any real-valued function, \gls{IS} is concerned with
estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
The posterior distributions of all but the simplest models' parameters fall
into this category. Suppose we have at hand another distribution $\eta$, called
the \defn{importance distribution}, from which we are able to sample. Define
the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
can express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
integral of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated at each of them by an empirical distribution. Since the remaining
quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
simple and elegant approach, the drawback is that the variance of the estimate
is proportional to the variance of the importance
weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
own is limited, since it depends on finding an importance distribution which is
similar to $\pi$, which we usually know very little about \textit{a priori}.

However, an ideal context for the use of \gls{IS} is when we want to sample
from a sequence of nested probability distributions $\pi_1, \ldots, \pi_n$.
By \defn{nested}, we mean that $\pi_{i+1}$ is defined on a space of dimension
$i+1$ and admits $\pi_i$ as a marginal. That is,
\[
  \pi_{i+1}(x_1, \ldots, x_{i+1}) = \pi_i(x_1, \ldots, x_i) f_{i+1}(x_1, \ldots, x_{i+1}),
\]
where $f_{i+1}$ is a function which yields a distribution when multiplied with
$\pi_i$. This situation may seem somewhat contrived, but it arises naturally
when trying to infer the hidden sequence of parameters of a stateful model. For
example, \textcite{doucet2001introduction} discuss the case when $\pi_i$ is the
posterior distribution over the first $i-1$ states of a \gls{HMM}, conditional
on the observed data up to time $i-i$. We assume that either $\pi_1$ is known
explicitly, or we have enough information about it to find an adequate
importance distribution. In the \gls{HMM} example, $\pi_1$ was taken to be the
prior distribution on the starting state.

@@

A linear regression is a model of the relationship between some data $\vec{x} =
(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
assumes that the outcomes are linearly related to the data, modulo some noise
introduced by, say, measurement errors and environmental fluctuations. It other
words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\[
  y_i = \beta x_i + \N(0, \sigma^2),
\]
where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
the model. If the $y_i$ are all independent, then for particular choice of
$\beta$ and $\sigma$, we can write down the probability density of observing
$\vec{y}$ given $\vec{x}$.
\[
  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
\]
The higher the value of this probability density, the more likely the
observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
natural criterion for choosing the parameters: we want to pick $\beta$ and
$\sigma$ which define a model where the probability of our observed data is as
high as possible. When performing such an optimization for fixed data, the
density function just written is also called the \defn{likelihood} of the
parameters,
\[
  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
\]
The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
\vec{x}, \vec{y})$ are called the \gls{ML} estimates.

@@

The main idea of \gls{ABC} is to replace the traditional likelihood-based
criteria for model ``goodness'' with one based on simulated data generated by
the model. Good models should be able to simulate data which closely resemble
reality. In \gls{ABC}, we are no longer interested in the posterior
distribution, but rather in the distribution of model parameters which produce
data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
level. The objective of \gls{ABC} is to recover the distribution\ldots

@@

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly}. Effectively, it comes down to the
approach described in the previous subsection of guessing parameter values
until one is close enough to the truth. More specifically, a possible set of
parameters $\theta$ is drawn from the prior distribution, and a simulated data
set $z$ is generated from the model with those parameters. If the distance
between the simulated data set and the real data, $\rho(\eta(y), \eta(z))$, is
small enough, then we accept $\theta$ as a sample from the posterior. This can
be repeated until as many samples as desired are obtained.
